---
title: "Reproducible Research: Peer Assessment 1"
output: 
  html_document:
    keep_md: true
---

```{r Set options and load R packages}
knitr::opts_chunk$set(warning=FALSE, fig.height=5, fig.width=8, cache=TRUE)
library(dplyr)
library(ggplot2)
```

## Loading and preprocessing the data
```{r Load and preprocessing the data}
activity <- read.csv("activity.csv")
activity$date <- as.Date(activity$date, "%Y-%m-%d")
summary(activity)
```


## What is mean total number of steps taken per day?
Below is a histogram showing the total number of steps taken per day:
  
```{r Histogram of the total number of steps taken per day}
tot_steps <- with(activity, aggregate(x=steps, by=list(date), FUN=sum, na.rm=TRUE))
names(tot_steps) <- c("date", "steps")
hist(x=tot_steps$steps, breaks=seq(0, 25000, by=1000), main="Histogram of total number of steps taken per day", xlab="Total number of steps taken per day")
```
  
The **mean** of the total number of steps taken per day is 
```{r Mean total step num}
mean(tot_steps$steps)
```
while the **median** is 
```{r Median total step num}
median(tot_steps$steps)
```


## What is the average daily activity pattern?
Below is a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days:

```{r Average daily activity pattern}
avg_act <- with(activity, aggregate(x=steps, by=list(interval), FUN=mean, na.rm=TRUE))
names(avg_act) <- c("interval", "mean_step_num")
plot(x=avg_act$interval, y=avg_act$mean_step_num, type="l", main="Number of steps taken on average across the duration", xlab="Interval", ylab="Average number of steps", lwd = 1.5)
```

The 5-minute interval that contains the maximum number of steps on average is
```{r Interval with max step num}
avg_act[which.max(avg_act$mean_step_num), ]$interval
```


## Imputing missing values
The total number of missing values in the dataset is
```{r total missing values num}
sum(is.na(activity$steps))
```

The missing values are filled in with the mean number of steps of the interval they belong to:
```{r Impute missing values}
filled_na <- avg_act$mean_step_num[match(activity$interval, avg_act$interval)]
```

Add in the imputed values to create a new dataset:
```{r New dataset with imputed values}
new_activity <- transform(activity, steps=ifelse(is.na(activity$steps), yes=filled_na, no=activity$steps))
new_tot_steps <- aggregate(new_activity$steps~new_activity$date, FUN=sum)
names(new_tot_steps) <- c("date", "steps")
```

The histogram of total number of steps taken per day is updated to be compared with the previous graph:
```{r New histogram of the total number of steps taken per day}
hist(new_tot_steps$steps, breaks=seq(0, 25000, by=1000), main="Histogram of total number of steps taken per day", xlab="Total number of steps taken per day", ylim=c(0, 20))
```

The new **mean** of the total number of steps taken per day is
```{r New mean total step num}
mean(new_tot_steps$steps)
```
while the new **median** is 
```{r New median total step num}
median(new_tot_steps$steps)
```

The values are different from what were seen previously. The imputation of missing data has increased the estimates of the total daily number of steps.


## Are there differences in activity patterns between weekdays and weekends?
A new factor variable 'day' is created to indicate if a given day is a weekday or weekend:
```{r Create variable to distinguish weekdays from weekends}
new_activity <- mutate(new_activity, day=weekdays(new_activity$date))
weekdays_list <- c('Monday', 'Tuesday', 'Wednesday', 'Thursday', 'Friday')
new_activity$day <- factor((weekdays(new_activity$date) %in% weekdays_list), 
         levels=c(TRUE, FALSE), labels=c('Weekdays', 'Weekends'))
```

To show the differences in activity patterns between weekdays and weekends, a panel of time series plots of the time interval and average number of steps taken, averaged across all weekdays or weekends, is created as below:
```{r Panel of time series plots}
act_days <- aggregate(steps~interval + day, data=new_activity, FUN=mean)
ggplot(data=act_days, aes(x=interval, y=steps, color=day)) + 
  geom_line() + 
  labs(title="Average daily steps by type of date", x="Interval", y="Average number of steps") + 
  facet_wrap(facets=~day, nrow=2, ncol=1)
```
